Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-05 sections/intro.tex
squish

The success and ubiquity of arrays is largely due to their simplicity. 
%
Since their introduction, multidimensional arrays have represented dense, rectilinear,
integer grids of points. 
%
By \textbf{dense}, we mean that indices are mapped to value via a simple formula relating multidimensional space to linear memory.
%
Consequently, dense arrays offer extensive compiler optimizations and many convenient interfaces.
%
%This
%simplicity enables extensive interoperability, convenience layers, and
%optimizations by breaking the abstraction barrier between array representation
%and array storage.  
%
Compilers understand dense array computations across many
programming constructs, such as for and while loops, breaks, parallelism,
caching, prefetching, multiple outputs, scatters, gathers, vectorization,
loop-carry-dependencies, and more. A myriad of optimizations have been developed for
dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
interchange.
%
However, while dense arrays are the easiest way to program for performance, the world is not all dense.

\begin{figure}
	\includegraphics[width=\linewidth]{Structures-examples.png}
    \vspace{-12pt}
    \caption{A few examples of matrix structures arising in practice}
    \vspace{-8pt}
\end{figure}
\kyle{In this paragraph, I wonder if it should be more "our world is full of structured data" which can be described by structured arrays. The "X describes Y" framing feels less fundamental than "Y is Z" like "images, masks, geometry, and transactions consist of long runs of identical values that can be compressed via run length encoding". This might be overly clunky though.}
Our world is full of structured arrays.
%
Sparse arrays (which store only nonzero elements) describe networks, databases, and simulations~\cite{abhyankarpetsc, bell2007lessons, mcauley2013hidden, balay2020petsc}.
%
Run-length encoding describes images and masks, geometry, and
databases (such as a list of transactions with the date field all the same)~\cite{shi2020column,golomb1966run}.
%
Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g. higher order FEMs) as well as in intermediate structures in many linear solvers (e.g. GMRES)~\cite{ded, saad2003iterative, o2009scientific}.
%
In the context of machine learning, combinations of sparse and blocked matrices are increasingly under consideration~\cite{dao2022monarch}.
%
Even complex operators can be expressed as structured arrays.
%
For example, a convolution with a filter can be expressed as a matrix multiplication
with the toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.

>>>>

The success and ubiquity of arrays is largely due to their simplicity. 
%
Since their introduction, multidimensional arrays have represented dense, rectilinear,
integer grids of points. 
%
By \textbf{dense}, we mean that indices are mapped to value via a simple formula relating multidimensional space to linear memory.
%
Consequently, dense arrays offer extensive compiler optimizations and many convenient interfaces.
%
%This
%simplicity enables extensive interoperability, convenience layers, and
%optimizations by breaking the abstraction barrier between array representation
%and array storage.  
%
Compilers understand dense computations across many
programming constructs, such as for and while loops, breaks, parallelism,
caching, prefetching, multiple outputs, scatters, gathers, vectorization,
loop-carry-dependencies, and more. A myriad of optimizations have been developed for
dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
interchange.
%
However, while dense arrays are the easiest way to program for performance, the world is not all dense.

\begin{figure}
	\includegraphics[width=\linewidth]{Structures-examples.png}
    \vspace{-12pt}
    \caption{A few examples of matrix structures arising in practice}
    \vspace{-8pt}
\end{figure}

Our world is full of structured data.
%
Sparse arrays (which store only nonzero elements) describe networks, databases, and simulations~\cite{abhyankarpetsc, bell2007lessons, mcauley2013hidden, balay2020petsc}.
%
Run-length encoding describes images and masks, geometry, and
databases (such as a list of transactions with the date field all the same)~\cite{shi2020column,golomb1966run}.
%
Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g. higher order FEMs) as well as in intermediate structures in many linear solvers (e.g. GMRES)~\cite{ded, saad2003iterative, o2009scientific}.
%
In the context of machine learning, combinations of sparse and blocked matrices are increasingly under consideration~\cite{dao2022monarch}.
%
Even complex operators can be expressed as structured arrays.
%
For example, a convolution with a filter can be expressed as a matrix multiplication
with the toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.
